Sketching for Density Estimation

### 2018-11-10 density.tex
some edit

\newcommand{\ip}[1]{\left \langle #1 \right \rangle}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\D}{\text{Density}}
\newcommand{\eps}{\epsilon}
%\usepackage{ntheorem}
%\usepackage{amsthm}

@@

\section{Problem Definition}
Given a collection of data points $X = x_1,\ldots, x_n$ in $\R^d$ the density of a point $y$ is defined as 
$$ \D(y) = \sum_{x \in X} K(y,x) $$
Here, $K$ is a \emph{positive definite kernel} function, typically based on the distance between $x,y$. The most frequent example is 
$$ K(x,y) = \exp(-\lambda \|x-y\|_2^2) $$
where $\lambda$ is a scaling parameter. What we discuss in what follows applies for any kernel, an depends on a bound on $K(x,x)$. For simplicity we assume that $K(x,x) \leq 1$ for all datapoints. Notice that for any kernel based on distance we have $K(x,x)=1$ exactly for all $x \in \R^d$.

Our objective is obtaining a core-set $z_1,\ldots, z_\ell \in \R^d$, $\ell \ll n$ along with weights $w_1,\ldots, w_\ell$ such that for any point $y \in \R^d$
$$ \left| \sum_{i=1}^\ell w_i \cdot K(z_i, y) - \D(y) \right| < \eps n $$
We are interested in achieving this in the streaming setting, while bounding the amount of memory used

@@

We continue to solve the following problem: Given $k$ vectors $z_1,\ldots,z_k$ in an inner product space, compute signs minimizing 
$$\| \sum_i s_i z_i \| $$
We first notice that for i.i.d uniform signs, it holds that
$$\E[\| \sum_i s_i z_i \|^2] = \E[\sum_{i,j} s_i s_j \ip{z_i, z_j}] = \sum_i \|z_i\|^2 $$
leading to a randomized approach. We use the method of expectation minimization to achieve this bound deterministically. Specifically, we set $s_i$ sequentially for $i$ while minimizing the expectation. We prove by induction on $i$ that for the fixed $s_j$'s up to and not including $i$,
$$\E_{s_i,\ldots,s_k}[\| \sum_j s_j z_j \|^2] \leq \sum_j \|z_j\|^2$$
For $i=1$ this is trivial. For $i>1$ let $m=\sum_{j=1}^{i-1} s_j z_j$.
$$ \E_{s_i,\ldots,s_k}[\| \sum_j s_j z_j \|^2] = \E[\| m + \sum_{j=i}^k s_j z_j \|^2] = \|m\|^2 + \sum_{j=i}^k \|z_j\|^2$$
It follows that by fixing $s_i$ deterministically,
$$ \E_{s_i,\ldots,s_k}[\| \sum_j s_j z_j \|^2] - \E_{s_{i+1},\ldots,s_k}[\| \sum_j s_j z_j \|^2] = \|m\|^2 + \|z_i\|^2 - \|m + s_i z_i\|^2  $$
so in order to make sure that 
$$ \E_{s_i,\ldots,s_k}[\| \sum_j s_j z_j \|^2] \geq \E_{s_{i+1},\ldots,s_k}[\| \sum_j s_j z_j \|^2] $$
we need
$$ 2\ip{m,s_i z_i} = \|m\|^2 + \|z_i\|^2 - \|m + s_i z_i\|^2 \geq 0 $$
We thus set 
$$ s_i = -\text{sign} \ip{m,z_i}$$

Let us now translate this to an algorithm w.r.t the $d$ dimensional points $x_1,\ldots, x_k$. The sign of $x_1$ is set arbitrarily as $s_1=1$. For $i>1$, we choose 
$$ s_i = -\text{sign} \left( \ip{\sum_{j<i} s_j \phi(x_j), \phi(x_i)} \right) = -\text{sign} \left( \sum_{j<i} s_j K(x_j, x_i) \right) $$
This guarantees that for any $y \in \R^d$,
$$ \sum_i s_i K(x_i, y) \leq \sqrt{ \sum_i \|\phi(x_i)\|^2 } = \sqrt{k} $$

@@

>>>>

\newcommand{\ip}[1]{\left \langle #1 \right \rangle}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\D}{\rho}
\newcommand{\eps}{\epsilon}
%\usepackage{ntheorem}
%\usepackage{amsthm}

@@

\section{Problem Definition}
Given a collection of data points $X = x_1,\ldots, x_n$ in $\R^d$ the density function $\rho: \R^d \rightarrow \R$ of a point $y$ is defined as 
$$ \rho(y) = \sum_{i=1}^{n} K(x_i,y) $$
Here, $K$ is a \emph{positive definite kernel} function, typically based on the distance between $x,y$. The most frequent examples include
$$ K(x,y) = \exp(- \|x-y\|_2^2/\sigma^2)\;\;\; K(x,y) = \exp(-\lambda \|x-y\|/\sigma) \; \mbox{and}\;\;\; K(x,y) = \frac{1}{1+\|x-y\|_2^2/\sigma^2)}$$
where $\sigma$ is a scaling parameter. What we discuss in what follows applies for any kernel, an depends on a bound on $K(x,x)$. For simplicity we assume that $K(x,x) \leq 1$ for all datapoints. Notice that for any kernel based on distance we have $K(x,x)=1$ exactly for all $x \in \R^d$.

Our objective is obtaining a core-set $z_1,\ldots, z_\ell \in \R^d$, $\ell \ll n$ along with weights $w_1,\ldots, w_\ell$ such that for any point $y \in \R^d$
$$ \left| \sum_{i=1}^\ell w_i \cdot K(z_i, y) - \D(y) \right| < \eps n $$
We are interested in achieving this in the streaming setting, while bounding the amount of memory used throughout the run of the algorithm.

@@

We continue to solve the following problem: Given $k$ vectors $z_1,\ldots,z_k$ in an inner product space, compute signs minimizing 
$$\| \sum_i s_i z_i \| $$
We first notice that for i.i.d.\ uniform signs, it holds that
$$\E[\| \sum_i s_i z_i \|^2] = \E[\sum_{i,j} s_i s_j \ip{z_i, z_j}] = \sum_i \|z_i\|^2 $$
leading to a randomized approach. By Markov with probability at least $3/4$ we have $\| \sum_i s_i z_i \| \le 2\sum_i \|z_i\|$.

We can also use the method of expectation minimization to achieve this bound deterministically.
Specifically, we will guaranty that $\| \sum_{j=1}^i s_j z_j \| \le \sum_{j=1}^i \|z_j\|^2$ for all $i$.
This is trivially true for $i=1$. 
For another value $i$ set $s_i = -\operatorname{sign} (\sum_{j=1}^{i-1}s_j \langle z_j, z_i \rangle)$ 
and  assume by induction that $\| \sum_{j=1}^{i-1} s_j z_j\|^2 \le \sum_{j=1}^{i-1} \|z_j\|^2$.

\begin{eqnarray*}
\| \sum_{j=1}^{i}s_j z_j\|^2 &=& \|\sum_{j=1}^{i-1}s_j z_j\|^2 + \|z_i\|^2 + \langle \sum_{j=1}^{i-1}s_j z_j, s_i z_i\rangle \\
&\le& \sum_{j=1}^{i-1} \|z_j\|^2 + \|z_i\|^2 + s_i \sum_{j=1}^{i-1}s_j \langle  z_j,  z_i\rangle \mbox{\;\;\; by the induction assumption}\\ 
&=& \sum_{j=1}^{i} \|z_j\|^2 - |\sum_{j=1}^{i-1}s_j \langle  z_j,  z_i\rangle| \le \sum_{j=1}^{i} \|z_j\|^2
\end{eqnarray*}

Let us now translate this to an algorithm w.r.t.\ the $d$ dimensional points $x_1,\ldots, x_k$. 
The sign of $x_1$ is set arbitrarily as $s_1=1$. For $i>1$, we choose 
$$ s_i = -\operatorname{sign} (\sum_{j=1}^{i-1}s_j \langle \phi(x_j), \phi(x_i) \rangle) = -\operatorname{sign} (\sum_{j=1}^{i-1}s_j  K(x_j, x_i)) $$
This guarantees that for any $y \in \R^d$,
$$ \sum_i s_i K(x_i, y) \leq \sqrt{ \sum_i \|\phi(x_i)\|^2 } = \sqrt{k} $$

Before we extend this to a complete algorithm, it is worth getting an intuition for the meaning of this result.

@@

Let us choose the value  and assume by induction that

%Specifically, we set $s_i$ sequentially for $i$ while minimizing the expectation. 
%We prove by induction on $i$ that for the fixed $s_j$'s up to and not including $i$,
%$$\E_{s_i,\ldots,s_k}[\| \sum_j s_j z_j \|^2] \leq \sum_j \|z_j\|^2$$
%For $i=1$ this is trivial. For $i>1$ let $v=\sum_{j=1}^{i-1} s_j z_j$.
%$$ \E_{s_i,\ldots,s_k}[\| \sum_{j=1}^k s_j z_j \|^2] = \E[\| v + \sum_{j=i}^k s_j z_j \|^2] = \|v\|^2 + \sum_{j=i}^k \|z_j\|^2$$
%It follows that by fixing $s_i$ deterministically,
%$$ \E_{s_i,\ldots,s_k}[\| \sum_j s_j z_j \|^2] - \E_{s_{i+1},\ldots,s_k}[\| \sum_j s_j z_j \|^2] = \|v\|^2 + \|z_i\|^2 - \|v + s_i z_i\|^2  $$
%so in order to make sure that 
%$$ \E_{s_i,\ldots,s_k}[\| \sum_j s_j z_j \|^2] \geq \E_{s_{i+1},\ldots,s_k}[\| \sum_j s_j z_j \|^2] $$
%we need
%$$ 2\ip{m,s_i z_i} = \|m\|^2 + \|z_i\|^2 - \|m + s_i z_i\|^2 \geq 0 $$
%We thus set 
%$$ s_i = -\text{sign} \ip{m,z_i}$$
